Towards open collaboration in biomedical and health research: An open collaborative scoping review

### 2025-02-10 doc/paper.qmd

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE)
library(stringr)
library(here)
source(here("R/openalex-search.R"))
search_terms_inline <- glue::glue("({search_terms()$title_search}[title]) AND {search_terms()$general_search}")
reviewed_records <- targets::tar_read(reviewed_article_records)

list_refs <- bib2df::bib2df(here("doc/reviewed-records.bib")) |>
  dplyr::filter(CATEGORY != "COMMENT") |>
  glue::glue_data("@{BIBTEXKEY}") |>
  str_c(collapse = ";") |>
  append("[", 0) |>
  append("]") |>
  str_c(collapse = "")
```

@@

We found `r nrow(reviewed_records)` source documents discussing open
collaboration in some way. These were `r list_refs`. Two of the
documents, while passed our abstract review, did not contain anything
relevant when we reviewed the full text [@Network2023a; @Yordanov2022],
so we did not continue with charting them.

### Practices of open collaboration

Across the papers found in our literature search, X out of 11 discuss
actual experiences on implementing open collaboration, albeit in
different organizational levels. These range from lab level
[@Turoman2022], center level [@Bush2022; @Network2023], multi-center
level [@ManyPrimates2021; @Network2023] and field level: psychology
[@Alessandroni2022], spinal cord injury [@TorresEspin2021] and
neuroimaging [@Niso2022]. All of these share in common the need for an
infrastructure that promotes open science and open collaboration via
starting building a community of collaborators or making tools (usually
online resources) that researchers can use to reach FAIR (Finndable,
Accessible, Interoperable and Reusable) goals while doing research.
Before delving in those, here we provide the specific needs for
implementing open collaboration in each organizational level, as well as
describing the differences in approaches among experiences between and
across levels (Figure 2).

#### Implementation of open collaboration practices at laboratory level

#### Implementation of open collaboration practices at center level

For center-level only one paper was identified [@Bush2022], providing
the example of a center of neuroimaging describing first how they
identified the needs for open collaboration, then the individual-level
practices that researchers should adapt and then the center-level
changes to promote reaching FAIR goals.

#### Implementation of open collaboration practices at multi-center level

Interestingly, the level of organization with most papers is the
multi-center level with two papers from ManyPrimates [@ManyPrimates2021;
@Primates2019] and one from the Reproducibility Network [@Network2023].
In this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, the Reproducibility Network
[@Network2023] describes a systematic review as an exploration of the
needs for 11 UKRN centers to implement open collaborative practives. In
this paper they specially focus on tools and resources among the 11
institutions and make clear the need for starting to utilize open
resources as a first step to adapt open collaborative practices. The
first paper from ManyPrimates [@Primates2019] describes the start of
another multi-center collaboration, albeit from different institutions.
It is interesting to observe that in this situation, the complex issue
is to start the collaboration itself, which was ignited thanks to a
symposium and an email chain among future collaborators. In contrast,
ManyPrimates could quickly develop an open collaborative structure with
relative ease in comparison to changing an already solid infrastructure,
like in the case of @Network2023.

#### Implementation of open collaboration practices at field level

>>>>

```{r setup}
#| include: false
knitr::opts_chunk$set(echo = FALSE)
library(targets)
library(readr)
library(dplyr)
library(glue)
library(stringr)
library(here)
source(here("R/openalex-search.R"))
search_terms_inline <- glue("({search_terms()$title_search}[title]) AND {search_terms()$general_search}")
reviewed_records <- tar_read(reviewed_article_records)

list_refs <- bib2df::bib2df(here("doc/reviewed-records.bib")) |>
  filter(CATEGORY != "COMMENT") |>
  glue_data("@{BIBTEXKEY}") |>
  str_c(collapse = ";") |>
  append("[", 0) |>
  append("]") |>
  str_c(collapse = "")
```

@@

```{r}
#| echo: false
n_papers <- tar_read(records_title_only) |>
  nrow()

n_papers_title_review <- tar_read(records_after_title_exclusion) |>
  nrow()

n_papers_excluded_title_review <- n_papers - n_papers_title_review

n_titles_selected <- tar_read(titles_selected) |>
  nrow()

n_paper_full_text <- tar_read(abstracts_selected) |>
  nrow()

n_papers_selected_for_review <- read_csv(here("data", "article_characteristics.csv")) |>
  nrow()

# get table of study characteristics
article_characteristics <- read_csv(here("data", "article_characteristics.csv"))

n_guides <- article_characteristics |>
  filter(paper_type == "guide") |>
  nrow()

n_examples <- article_characteristics |>
  filter(paper_type == "example") |>
  nrow()

n_surveys <- article_characteristics |>
  filter(paper_type == "survey") |>
  nrow()

n_org <- article_characteristics |>
  filter(str_detect(themes, "organization")) |>
  nrow()

n_tools <- article_characteristics |>
  filter(str_detect(themes, "tools")) |>
  nrow()

n_education <- article_characteristics |>
  filter(str_detect(themes, "education")) |>
  nrow()

n_researcher <- article_characteristics |>
  filter(relevant_for == "researcher") |>
  nrow()

n_research_group <- article_characteristics |>
  filter(relevant_for == "research group") |>
  nrow()

n_research_center <- article_characteristics |>
  filter(relevant_for == "research center") |>
  nrow()

n_research_centers <- article_characteristics |>
  filter(relevant_for == "multiple centers") |>
  nrow()

n_research_field <- article_characteristics |>
  filter(relevant_for == "research field") |>
  nrow()
```

Our search yielded a total of `r n_papers` papers from which we filtered
`r n_papers_excluded_title_review` based on key words. Three reviewers
independently screened for the remaining `r n_papers_title_review`
titles from which we screened `r n_titles_selected` abstracts. Of these,
we assessed `r n_paper_full_text` full-text papers and we included
`r n_papers_selected_for_review` in the scoping review. These were
`r list_refs`. Two of the documents, while they passed our abstract
review, did not contain anything relevant when we reviewed the full text
[@Network2023a; @Yordanov2022], so we did not continue with charting
them.

The identified papers included `r n_guides` guides, `r n_examples`
examples and `r n_surveys` surveys.

`r n_org` of the `r n_papers_selected_for_review` papers included
recommendations for organizational practices to improve open
collaboration, `r n_tools` mentioned specific tools used for open
collaboration and `r n_education` highlighted educational needs. The
identified papers discussed open collaboration on different
organizational levels. `r n_researcher` were mostly directed at the
individual researcher, `r n_research_group` at a research group,
`r n_research_center` at a research center, `r n_research_centers` at
several research centers and `r n_research_field` at entire research
fields.

### Organizational practices of open collaboration

A common theme across organizational levels was the need for
infrastructure that promotes open collaboration, building a community of
collaborators and using online tools in accordance with the FAIR
(Findable, Accessible, Interoperable and Reusable) principles.

individual level - add text

laboratory/research group level - add text

For center-level one paper described how they transitioned their
neuroimaging research center toward more open and collaborative
practices. Their central focus was on creating more open code and data
to improve open collaboration [@Bush2022]. First, they used open-source
code, then standardized data formats, developed analysis pipelines,
shared data openly and developed protocols for data documentation. While
the organizational transition to improve open collaboration was
described, little was mentioned about how this has improved
collaborative work at the research center.

Interestingly, the level of organization with most papers was the
multi-center level with two papers from ManyPrimates [@Primates2019;
@ManyPrimates2021] and the Reproducibility Network [@Network2023]. In
this organization level the most urgent needs are: 1) identifying the
needs for open collaboration and 2) creating an infrastructure that
allows those needs to be satisfied. Indeed, Grange et al describes a
systematic review as an exploration of the needs for 11 UKRN centers to
implement open collaborative practives. In this paper they specially
focus on tools and resources among the 11 institutions and make clear
the need for starting to utilize open resources as a first step to adapt
open collaborative practices. The first paper from ManyPrimates
[@Primates2019] describes the start of another multi-center
collaboration, albeit from different institutions. It is interesting to
observe that in this situation, the complex issue is to start the
collaboration itself, which was ignited thanks to a symposium and an
email chain among future collaborators. In contrast, ManyPrimates could
quickly develop an open collaborative structure with relative ease in
comparison to changing an already solid infrastructure, like in the case
of @Network2023.

research field level

Here, we provide the specific needs for implementing open collaboration
in each organizational level, as well as describing the differences in
approaches among experiences between and across levels (Figure 2).
